---
title: "Machine Learning Final Project - Predicting Song Success"
date: "December 11, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
```

## Contributers : 
* Anna Zagoren : azagoren@wesleyan.edu
* Erin Rose : erose@wesleyan.edu
* Emily Leff : eleff@wesleyan.edu

### **Importing files **
```{r}
# csv version of Million Song Subset came from:
# https://raw.githubusercontent.com/Vatshayan/Song-Classification/master/music.csv
# which we saved as "music.csv"

msd <- read.csv("music.csv")

# Billboard top 100 data came from:
# https://data.world/kcmillersean/billboard-hot-100-1958-2017
# which we saved as "billboard.csv"

billboard <- read.csv("billboard.csv")
```

### **Data Preprocessing**
```{r}
library(dplyr)

billboard %>%
  group_by(Song, Performer)%>%
  mutate(WeeksonChart=sum(Instance)) -> billboard

billboard_agg <- distinct(billboard, Song, Performer, WeeksonChart)

colnames(billboard_agg) <- c("title", "artist.name", "weeksonchart")

data <- left_join(msd, billboard_agg, by = c("title","artist.name"))

data$artist.name <- factor(data$artist.name)
data$title <- factor(data$title)
data[["weeksonchart"]][is.na(data[["weeksonchart"]])] <- 0
data$on_billboard <- factor(ifelse(data$weeksonchart>0, "yes", "no"))

# Data Removals
data <- data %>%
  select(-c(artist_mbtags, artist_mbtags_count, location, longitude,
            latitude, song.id, terms, terms_freq, year, song.hotttnesss))
data2 <- na.omit(data)

library(caret)
set.seed(1234)
index <- createDataPartition(data2$on_billboard, p=0.85, list=FALSE) 
train <- data2[index, ]
test  <- data2[-index, ]
```

### **Machine Learning Approaches**

* Classification Tree:
* Random Forest:
* Lasso Regression:
* Random forest using SMOTE sampling:

### **Results**

#### Classification Tree:
```{r}
# set train control
trctrl <- trainControl(method="cv", number=10,
                       summaryFunction=twoClassSummary,
                       classProbs=TRUE)

# build model
set.seed(1234)
ctree_fit <- train(on_billboard ~. -artist.id -artist.name -release.id -release.name -title -weeksonchart -similar, 
                   data = train, 
                   method = "rpart",
                   trControl=trctrl,
                   metric = "ROC",
                   tuneLength = 10)
ctree_fit

# graph classification tree
library(rattle)
fancyRpartPlot(ctree_fit$finalModel, tweak=1.4)

# evaluate on test data
pred <- predict(ctree_fit, test)
confusionMatrix(pred, test$on_billboard, positive="yes")

```

#### Random Forest:
```{r}
library(randomForest)

# build model
set.seed(1234)
fit.forest <- randomForest(on_billboard ~ . -artist.id -artist.name -release.id -release.name -title -weeksonchart -similar,
                           data=train,
                           na.action=na.roughfix,
                           ntree=1000,
                           importance=TRUE)

fit.forest

# variable importance graph
varImpPlot(fit.forest, type=2, main="Variable Importance")

# evaluate on test data
pred <- predict(fit.forest, test)
# library(caret)
confusionMatrix(pred, test$on_billboard, positive="yes")
```

#### Lasso Regression:
```{r}
require(glmnet)
x <- model.matrix(on_billboard ~ ., train)

# dependent variable
y <- train$on_billboard

# specify the lamba values to investigate
grid <- 10^seq(8, -4, length=250)

model.lasso <- glmnet(x, y, family="binomial")
plot(model.lasso, xvar="lambda", main="Lasso")

# 10-fold cross validation of lambda values on MSE
set.seed(1234)
cv.lambda <- cv.glmnet(x, y, family="binomial")
plot(cv.lambda)
bestlam <- cv.lambda$lambda.min

# evaluate model on test data 
x <- model.matrix(on_billboard ~ ., test)

pred <- predict(model.lasso, s=bestlam, newx=x, type="class")

pred <- pred[,1]
pred <- as.character(pred)
pred <- factor(pred, levels= c('no', 'yes'))
actual<- ifelse(test$on_billboard == "yes", "yes", "no")
actual <- factor(actual, levels= c('no', 'yes'))
confusionMatrix(actual, pred, positive="yes")
```

#### Random forest using SMOTE sampling:
```{r}
library(DMwR)

train <- train %>%
  select(-c(artist.id, artist.name, release.id, release.name, title, weeksonchart, similar))

test <- test %>%
  select(-c(artist.id, artist.name, release.id, release.name, title, weeksonchart, similar))

ctr <- trainControl(method = "repeatedcv", 
                    number = 10, 
                    repeats = 10, 
                    verboseIter = FALSE,
                    sampling = "smote")

set.seed(1234)
fit <- train(on_billboard ~ .,
             data = train,
             method = "rf",
             preProcess = c("scale", "center"),
             trControl = ctr)

pred <- predict(fit,  newdata = test)

confusionMatrix(pred, test$on_billboard, positive = "yes")
postResample(pred = pred,
             obs = test$on_billboard)
```


